---
title: Customize and Query Compound Annotation Database
author: "Authors: Yuzhu Duan, Dan Evans, Jamison McCorrison, Nicholas Schork, Kevin Horan, Austin Leong, Siddharth Sai and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{customCompoundDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
fontsize: 14pt
bibliography: bibtex.bib
---

<style>
pre code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('customCompoundDB.Rmd',
              c('BiocStyle::html_document', 'pdf_document'));
              knitr::knit('customCompoundDB.Rmd', tangle=TRUE)"
-->

# Introduction

This software package is the customization and query interface for the
annotation SQLite database from the corresponding
[compoundCollectionData](https://github.com/yduan004/compoundCollectionData) package.
It provides utilities to query the compound annotations from 
[DrugAge](https://genomics.senescence.info/drugs/), 
[DrugBank](https://www.drugbank.ca/),
[CMAP02](https://portals.broadinstitute.org/cmap/),
and [LINCS](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742) 
resources by providing ChEMBL ids of the query compounds.
It also supports adding custom compound annotations to the annotation SQLite
database.

# Installation and Loading
As Bioconductor package `customCompoundDB` can be installed with the 
`BiocManager::install()` function.
```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("customCompoundDB")
BiocManager::install("yduan004/customCompoundDB", build_vignettes=TRUE)  # Installs from github
```

Next the package needs to be loaded into a user's R session.
```{r load, eval=TRUE}
library(customCompoundDB)
library(help = "customCompoundDB")  # Lists package info
vignette("customCompoundDB")  # Opens vignette
```

# Annotation Database
The helper package `compoundCollectionData` provides access to the pre-built 
SQLite database that are stored on Bioconductor's
`AnnotationHub`. Users can download this database and get its path as follows.

```{r download_db, eval=TRUE, message=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
annotdb <- ah[["AH79563"]]
```

# Custom Annotations
## Add Custom Annotation
The following shows functions used to add user's customized compound annotation 
tables to the annotation SQLite database in the 
`compoundCollectionData` package. In this case, users need to know the 
corresponding ChEMBL ids of the added compounds. The added annotation table 
should contain the `chembl_id` column.
```{r custom, eval=TRUE}
chembl_id <- c("CHEMBL1000309", "CHEMBL100014", "CHEMBL10",
               "CHEMBL100", "CHEMBL1000", NA)
annot_tb <- data.frame(compound_name=paste0("name", 1:6),
        chembl_id=chembl_id,
        feature1=paste0("f", 1:6),
        feature2=rnorm(6))
addCustomAnnot(annot_tb, annot_name="mycustom")
```

`annot_tb` is an R data.frame object representing the custom annotation table, 
Note, it should contains a column named as `chembl_id` representing the ChEMBL 
ids of the added compounds. `annot_name` is a user defined name of the 
annotation table.

## Delete
The following shows the R code used to delete a custom annotation resource
by providing its name.
```{r del, eval=FALSE, message=FALSE}
deleteAnnot("mycustom")
```

## List Existing Annotations
The following function lists the available annotation resources in the SQLite 
annotation database
```{r list, eval=TRUE, message=FALSE}
listAnnot()
```

## Set to Default
The following function sets the annotation SQLite database to the default one 
by deleting the existing one and re-downloading from AnnotationHub.
```{r default, eval=FALSE}
defaultAnnot()
```

# Query Annotation DB
The following function can be used to query compound annotations from the default
resources as well as the custom resources stored in the SQLite annotation
database. The default annotation resources are `DrugAge`, `DrugBank`, `CMAP02`
and `LINCS`. Detailed description of this SQLite database is available at the
[vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/compoundCollectionData/inst/doc/compoundCollectionData.html) 
of the `compoundCollectionData` package. Users customized compound 
annotations could be added/deleted as described above. 

The input of the query function is a set of ChEMBL IDs, it returns a data.frame 
storing annotations of the input compounds from the selected annotation 
resources defined by the \code{annot} argument.
```{r query, eval=TRUE, message=FALSE}
query_id <- c("CHEMBL1064", "CHEMBL10", "CHEMBL113", "CHEMBL1004", "CHEMBL31574")
annot_res <- queryAnnotDB(query_id, annot=c("DrugAge", "LINCS"))
annot_res
# query added custom annotation
annot_res2 <- queryAnnotDB(query_id, annot=c("LINCS", "mycustom"))
annot_res2
```
